### 2020-10-15 submission/manuscript.Rmd
Create figure showing quality of singletons

**The nature of singletons.** Removal of rare ASVs is commonly justified as a method of removing ASVs that are artifacts. If such ASVs are artifacts, then one would expect the number of singleton ASVs to accumulate with sequencing depth. Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`). In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in the sample was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`). This showed that with additional sequencing, the probability of seeing rare ASVs multiple times was greater than the probability of generating an artifact. This suggests that the rare ASVs are not as likely to be artifacts as previously thought. Furthermore, if rare ASVs were artifacts, then one would not expect to find them in other samples from the same dataset. In fact, singleton ASVs from samples with fewer sequences were often found in samples with more sequences. At least 50% of the singleton ASVs found in the samples from the `r oxford_comma(high_coverage)` datasets were found in another sample from the same dataset. Considering the likelihood of finding an ASV duplicated in another sample is confounded by the number of samples and inter-sample diversity, the high coverage of singleton ASVs in these datasets was remarkable. The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively. The negative correlation indicated that the singleton ASVs in the smaller samples were more likely to be covered by ASVs in the larger samples. Among the three datasets without a significant correlation (Spearman correlation, P>0.05), the marine and soil datasets had the fewest samples in our collection and the stream dataset already had a high level of coverage regardless of the number of sequences. Contrary to the common motivation for removing rare ASVs, these results indicate that this practice disproportionately impacts samples with fewer sequences and likely removes a large number of non-artifact ASVs.

@@

**How do communities change when rare ASVs are removed?** Removing rare ASVs will reduce the number of observed ASVs and proportionally increase the relative abundance of the remaining ASVs. The result was expected to be a loss of information contained within each sample. To quantify the effect of removing rare ASVs on the information contained within each sample, I varied the minimum abundance threshold to simulate removing ASVs of varying rarity from each sample. The number of ASVs in each sample decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure 1). Similarly, the Shannon diversity decreased by between `r oxford_comma(asv_h_two, 1)`% when removing ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing ASVs that appeared ten or fewer times from each sample (Figure 1). Next, I assigned the ASVs to OTUs to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies. Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the number of observed OTUs and the diversity decreased relative to the original community (**Figure 1**). In contrast to the number of ASVs or OTUs and their diversity, the Kullback–Leibler divergence compares the relative abundance of each ASV or OTU between representations of the community. I calculated the Kullback–Leibler divergence from the full to pruned communities when rare ASVs were removed. As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure 1). The relative loss of information was generally lower for OTUs than than it was for ASVs. Removing rare ASVs, regardless of abundance threshold, had profound impacts on the representation of the communities.

>>>>

**The nature of singletons.** Removal of rare ASVs is commonly justified as a method of removing ASVs that are artifacts. If such ASVs are artifacts, then one would expect the number of singleton ASVs to accumulate with sequencing depth. Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`). In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in the sample was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figure 1A). This showed that with additional sequencing, the probability of seeing rare ASVs multiple times was greater than the probability of generating an artifact. This suggests that the rare ASVs are not as likely to be artifacts as previously thought. Furthermore, if rare ASVs were artifacts, then one would not expect to find them in other samples from the same dataset. In fact, singleton ASVs from samples with fewer sequences were often found in samples with more sequences. At least 50% of the singleton ASVs found in the samples from the `r oxford_comma(high_coverage)` datasets were found in another sample from the same dataset (Figure 1B). Considering the likelihood of finding an ASV duplicated in another sample is confounded by the number of samples and inter-sample diversity, the high coverage of singleton ASVs in these datasets was remarkable. The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively (Figure 1C). The negative correlation indicated that the singleton ASVs in the smaller samples were more likely to be covered by ASVs in the larger samples. Among the three datasets without a significant correlation (Spearman correlation, P>0.05), the marine and soil datasets had the fewest samples in our collection and the stream dataset already had a high level of coverage regardless of the number of sequences. Contrary to the common motivation for removing rare ASVs, these results indicate that this practice disproportionately impacts samples with fewer sequences and likely removes a large number of non-artifact ASVs.

@@

**How do communities change when rare ASVs are removed?** Removing rare ASVs will reduce the number of observed ASVs and proportionally increase the relative abundance of the remaining ASVs. The result was expected to be a loss of information contained within each sample. To quantify the effect of removing rare ASVs on the information contained within each sample, I varied the minimum abundance threshold to simulate removing ASVs of varying rarity from each sample. The number of ASVs in each sample decreased by between `r oxford_comma(asv_s_two, 1)`% when removing those ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing those that appeared ten or fewer times from each sample (Figure X). Similarly, the Shannon diversity decreased by between `r oxford_comma(asv_h_two, 1)`% when removing ASVs that only appeared once and by between `r oxford_comma(asv_s_eleven, 1)`% when removing ASVs that appeared ten or fewer times from each sample (Figure X). Next, I assigned the ASVs to OTUs to assess the impact of removing rare ASVs on higher level taxonomic groupings that are commonly used in microbial ecology studies. Although pooling similar ASVs into OTUs reduced the impact of removing the rare ASVs relative to the ASV-based analysis, the minimum abundance threshold still decreased the number of observed OTUs and the diversity decreased relative to the original community (**Figure 1**). In contrast to the number of ASVs or OTUs and their diversity, the Kullback–Leibler divergence compares the relative abundance of each ASV or OTU between representations of the community. I calculated the Kullback–Leibler divergence from the full to pruned communities when rare ASVs were removed. As the threshold for removing ASVs increased, the amount of information lost also increased for both ASVs and OTUs (Figure X). The relative loss of information was generally lower for OTUs than than it was for ASVs. Removing rare ASVs, regardless of abundance threshold, had profound impacts on the representation of the communities.
